The ANthropological Notation Ontology (ANNO): A core ontology for annotating human bones and deriving phenotypes

### 2024-01-25 paper.tex
MLH Changes 25-01-23, 07:41

\todo{Hier ist die Schnittstelle zwischen dem neuen Text aus Mittweida oben und dem existierenden theoretischen Hintergrund unten, eventuell muss das noch jemand glattbügeln.}
%\paragraph{Theoretical Background}
Ontologies are theories about the kinds of objects, the properties of objects, and the relations between objects in a knowledge domain.
On the one hand, these are controlled vocabularies, but on the other hand, ontologies are also conceptualizations that the vocabulary terms are intended to capture~\citep{whatareontologies}.
Researchers in many areas recognized the need for ontologies to clearly define specialized vocabularies for these domains~\citep{evaluatingdomain}.
The success of using ontologies to annotate biological data can also be confirmed by multiple examples~\citep{biomedicalontologies}.
It was therefore an easy decision for us to use an ontology for modeling the anthropological domain.

@@

\todo{new content from MH:}
The ontology was created to be used in conjunction with the anthropological 3D editor \aw{}.
AW3D allows users to examine digitized bone material which can be created using a pipeline that was designed specifically for creating 3D models of bones for use in anthropological morphological and osteometric research and examination.
To allow such an examination the software can import and render these 3D models at runtime and provides a wide suite of tools for the annotation and measuring of bone material.
The ontology was integrated into the software to better meet the use cases of AW3D.
These use cases include the annotation of bone material through markings in 3D space on the bone models, including point, line, and surface markings.
They also involve the measuring of bones by providing line, circumference, and angle measurement tools for use in osteometrical contexts.
To improve the quality of the annotations and measurements additional information such as alternative titles and descriptions, and others can be input by text.
Moreover, the software offers the capability to display either the entire skeleton or specific parts, thereby enhancing the examination context.
This feature enables users to not only focus on the specific bone they are studying but also easily access adjacent or related bones for a more comprehensive understanding.
These use cases are already covered by AW3D itself but are improved by an integration of ANNO into the software.
Additional use cases that are only achievable through this integration include the automatic derivation of bone and skeletal phenotypes using mathematical functions available in the ontology and the provision of anthropological, anatomical and osteometrical knowledge for users through ANNO.

@@

\todo{existing content:}
\aw{}, see \cref{fig:aw}, is a German-language tool that combines user-friendly techniques of photogrammetry with insights from user experience research and knowledge from game development and offers a procedure to create 3D skeletal models serving as digital twins which can subsequently be examined virtually.
This facilitates anthropological work to be location-independent and parallel without exercising wear and tear on the skeletal material.
The examination can be performed as often as desired, even if the skeletal individuals or collections are not available at the institute or have already been reburied.
%
We use our ontological architecture (top-level ontology, domain core ontology, domain specific ontology) to integrate ANNO into \aw{} based on the three-ontology method~\cite{threeontologymethod}.

>>>>

\todo{Hier ist die Schnittstelle zwischen dem neuen Text aus Mittweida oben und dem existierenden theoretischen Hintergrund unten, eventuell muss das noch jemand glattbügeln.}

%\paragraph{Theoretical Background}
Ontologies are theories about the kinds of objects, the properties of objects, and the relations between objects in a knowledge domain.
On the one hand, these are controlled vocabularies, but on the other hand, ontologies are also conceptualizations that the vocabulary terms are intended to capture~\citep{whatareontologies}.
Researchers in many areas recognized the need for ontologies to clearly define specialized vocabularies for these domains~\citep{evaluatingdomain}.
The success of using ontologies to annotate biological data can also be confirmed by multiple examples~\citep{biomedicalontologies}.
It was therefore an easy decision for us to use an ontology for modeling the anthropological domain.

@@

\aw{}, see \cref{fig:aw}, is a German-language tool that combines user-friendly techniques of photogrammetry with insights from user experience research and knowledge from game development. 
It enables users to virtually examine digitized bone material, which can be created using a procedure designed for generating 3D models of bones. These models serve as digital twins in anthropological, morphological, and osteometric research and examination. To facilitate such examinations, the software can import and render these 3D models at runtime and provides a comprehensive suite of tools for annotating and measuring bone material.
This facilitates anthropological work to be location-independent and parallel without exercising wear and tear on the skeletal material. The examination can be performed as often as desired, even if the skeletal individuals or collections are not available at the institute or have already been reburied.
The ANNO ontology was created to be used in conjunction with the anthropological 3D editor AW3D \aw{} and was integrated into the software to better meet the use cases of the program. These use cases include the annotation of bone material through markings in 3D space on the bone models, including point, line, and surface markings. They also involve the measuring of bones by providing line, circumference, and angle measurement tools for use in osteometrical contexts.
To improve the quality of the annotations and measurements additional information such as alternative titles and descriptions, and others can be input by text.
Moreover, the software offers the capability to display either the entire skeleton or specific parts, thereby enhancing the examination context .This feature enables users to not only focus on the specific bone they are studying but also easily access adjacent or related bones for a more comprehensive understanding.
These use cases are already covered by AW3D itself but are improved by an integration of ANNO into the software. Additional use cases that are only achievable through this integration include the automatic derivation of bone and skeletal phenotypes using mathematical functions available in the ontology and the provision of anthropological, anatomical and osteometrical knowledge for users through ANNO.

@@

%
We use our ontological architecture (top-level ontology, domain core ontology, domain specific ontology) to integrate ANNO into \aw{} based on the three-ontology method~\cite{threeontologymethod}.
